Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-23 paper.tex
Tidy up intro to xlins

\section{Results}
\subsection{Inferred ARGs}
The goals of this preliminary study are to illustrate the utility of
\texttt{sc2ts}, and to closely study the properties of the
inferred ARGs to inform subsequent development. 
We work with a 
representative subset of the available data, limited to
inferences that can be performed on a single server in
a few weeks (see below for further details on timings and
computer hardware used). The cut-off dates for sampling were chosen
arbitrarily.
We inferred two ARGs, which we refer to as the
``Wide'' and ``Long'' ARGs throughout.
The Wide ARG is densely sampled but time-limited and
includes all 1.27 million sample genomes  with a
maximum sampling delay of 30 days
(see Section~\ref{sec:filtering_time_travellers})
up to June 30, 2021.
The Long ARG is more sparsely sampled but
includes data for an additional year (June 30, 2022).
For the long ARG, we use at most 1,000 randomly subsampled
genomes per day from those meeting the maximum sampling delay criterion.

@@

In this section we focus on the detailed genealogy of samples that have been
previously identified as recombinants, i.e. given a Pango designation that
starts with an ``X''. As the Wide ARG was only sampled until mid 2021,
the majority of these Pango X lineages are found only in the Long ARG.

\subsubsection{Long ARG}
%% high level summary of the pango recombinant lineages and samples in the Long ARG
The Long ARG (subsampled to mid-2022) contains 33
Nextclade-designated Pango recombinant lineages comprising 749 samples
(711 when singleton recombinants are removed). In
contrast, GISAID designates 38 Pango recombinant lineages in the long ARG,
comprising only 515 samples. In the analyses below, we mainly focus on the
larger number of Nextclade designated samples.

Of the 749 Nextclade-designated recombinants, we infer that all but two descend
from a recombination node. These two samples are designated XP, and a likely
explanation for the absence of a corresponding recombination event in the long ARG
is that the characteristic multibase deletion for XP
(\url{https://github.com/cov-lineages/pango-designation/issues/481}) is masked
during preprocessing. An improved masking pipeline is an important avenue for
future development (see the Discussion).

38 of the remaining X designated samples are singleton recombinants, descending
from recombination nodes that are ancestral only to that sample (10 are XZ; 6
are XE;  3 each from XN and XK; 2 each from XC, XS, XV, XQ and XAB; and 1
sample from each of XB, XM, XJ, XAF, XAH and XAJ). Such samples are likely to
be enriched for sequencing errors and lineage designation artefacts, and so we
exclude them from further analysis in this section.

The most recent recombination node for 79 samples (XN: 53, XZ: 16, XAJ: 6, XE:
1, XAD: 1, XAH: 1, XAK: 1) is the same. This node, caused by strain Denmark/DCGC-281594/2021,
is ancestral to more than 127K samples, most of which are not designated as
recombinants by any other method. It is highly likely to be a spuriously inferred recombination event.
Understanding the origin of this and other similar nodes is an important
direction for future work (see Discussion). For simplicity, we exclude these
samples from further analyses in this section.

The remaining remaining 630 Pango recombinant samples trace back to 50 different
most recent recombination nodes, representing 31
Nextclade-designated Pango lineages. These are summarised
in Supplementary Table~\ref{tab:pango-recombinants}, and fall
into three classes: single origin, multiple origin,
and multiple nested origins.

% The sequences of 28 Pango recombinants trace back
% to some recombinant node(s); however, the sequences of the five Pango
% % jk: really? I thought it was just XP that wasn't inferred to be a
% % recombinant.
% recombinants XP, XU, XAA, XAG, and XAK do not.
% A likely explanation for the XP
% recombinant sequences not having a recombinant origin in the long ARG is that
% its characteristic multibase deletion
% (\url{https://github.com/cov-lineages/pango-designation/issues/481}) was masked
% during preprocessing and therefore was not incorporated into the HMM.
% % Eh?
% Upon tracing the origin of XAG through the ARG, XAG might have been caused by
% incorrect imputation of its parent nodes(?). XAK likely does not appear to have
% a recombinant origin because of random subsampling, but it may with a more
% complete tree.

% [FIXME this isn't quite right - not all the sequences for a given pango
% lineage trace back to the same recombination event, so won't have the same
% intervals.]
% The mosaic genome structure of all the 32 Pango recombinant lineages is concordant with
% the structure determined by the Pango Lineage Designation Committee
% (Supplementary Table 3). For example, the genome of XD (also known as
% ``Deltacron'') is inferred to consist of B.1.617.2 (Delta) on the 5’ end, BA.1.17
% (Omicron) in the middle, and B.1.617.2 (Delta) on the 3’ end.

% For the long ARG, we examine the lineages in increasing complexity of the
% inferred evolutionary relationships between samples assigned to Pango
% X lineages.

% - Recombinant Pango lineages
%     How well do we recover recombinant Pango lineages? Work through the
% examples in increasing compexity, using headings to split things up if
% necessary. (i.e., Single origin, Multiple origin, Interellated lineages). We'd
% need to be clear about what these terms mean, of course. We could include
% lineage graphs for all of these examples?

% OLD TEXT

% The sequences of 14 Pango recombinants trace back to a single recombination
% node. The sequences of the remaining 18 Pango recombinants trace back to
% multiple recombination nodes, however. This may be a result of undersampling in
% the long ARG, which was built by randomly subsampling at most 1,000 sequences
% from each daily batch. Having extra recombination nodes in the long ARG
% appeared to be a more parsimonious explanation for the 19 Pango recombinants
% than having a single recombination node with many immediately ancestral
% mutations (including recurrent mutations and reversions), despite that the
% recombinants may indeed trace to a single origin.

@@

\caption{\label{fig:pango-simple-origin-graph} Non-nested
Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
imputed Pango status. Genomic regions inherited by the recombination node are shown
corresponding to the maximum inferred breakpoint position from the HMM process.
(B) Equivalent plot for the 17 XAG samples in the long ARG, with abbreviated labelling.
(C) Equivalent plot for both origination events involving the four XD samples in the long ARG. Details of the mutations and sample node identities for all three plots are provided in supplementary Figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}.
} \end{figure}

@@

It is clear from the XA subgraph that all the samples labelled XA by Nextclade trace to a
single originating recombination node, whose genome is a composite of a B.1.177.18 lineage
on the left of the genome and a B.1.1.7 lineage on the right. In the subgraph we show
the maximum genomic position for the recombination breakpoint, here at 22227 basepairs
(corresponding to a breakpoint strictly less than 22228, see Table~\ref{tab:jackson})

@@

\subsubsection{Wide ARG}
%% Quick overview of the wide ARG.
Since the Wide ARG is restricted to data collected prior
to mid-2021, it contains samples from only three Pango-designated recombinant
lineages: XA, XB, and XC. Both Nextclade and GISAID designate 44
samples as XA, while 237 samples are designated
as XB by Nextclade (231 by GISAID), and 6 as XC by Nextclade (none by GISAID).
After removing singleton recombinants, XA numbers remain unchanged, but XB is
reduced to 235 Nextclade-designated samples (229 GISAID) and XC is reduced to 4
(none in the GISAID designations). We correctly identify that all samples
designated as XA, XB, or XC by any method have one or more recombination nodes
in their ancestry.

>>>>

\section{Results}
\subsection{Inferred ARGs}
The goals of this preliminary study are to illustrate the utility of
\texttt{sc2ts}, and to closely study the properties of the
inferred ARGs to inform subsequent development.
We work with a
representative subset of the available data, limited to
inferences that can be performed on a single server in
a few weeks (see below for further details on timings and
computer hardware used). The cut-off dates for sampling were chosen
arbitrarily.
We inferred two ARGs, which we refer to as the
``Wide'' and ``Long'' ARGs throughout.
The Wide ARG is densely sampled but time-limited, and
includes all 1.27 million sample genomes collected up to June 30, 2021
with a maximum delay between sampling and submission dates of 30 days
(see Section~\ref{sec:filtering_time_travellers}).
% The Long ARG is more sparsely sampled but
% includes data for an additional year (June 30, 2022).
% For the Long ARG, we use at most 1,000 randomly subsampled
% genomes per day from those meeting the maximum sampling delay criterion.
For the Long ARG, we randomly sub-sample a maximum of 1,000 genomes per
day (again restricting the delay between sampling and submission to 30 days),
and include an additional year worth of samples (to June 30, 2022).

@@

In this section we focus on the detailed genealogy of samples that have been
previously identified as recombinants, i.e. given a Pango designation that
starts with an ``X''. We primarily focus on the Long ARG in this section
since it contains many more recombinants, briefly summarising the status
of Pango X lineages for the Wide ARG in Section~\ref{sec-pango_x_wide_arg}.

When using Nextclade Pango lineage designations, the
Long ARG contains 749 samples from 33 recombinant lineages; this
reduces toe 711 and XX when singleton recombinants (i.e., recombination
nodes subtending only one sample) are removed.
In contrast, when using GISAID Pango lineage assignments (for the same
data) we have 515 samples from 38 Pango recombinant lineages,
indicating that there is substantial uncertainty in the assignment of
Pango X lineages to samples.

Of the 749 Nextclade-designated recombinant samples, all but two descend
from a recombination node in the Long ARG.
These two samples are designated XP, and a likely
explanation for the absence of a corresponding recombination event in the long ARG
is that the characteristic multibase deletion for XP
(\url{https://github.com/cov-lineages/pango-designation/issues/481}) is masked
during preprocessing (See~\ref{sec:data_preprocessing} for details and potential
improvements). 38 of the remaining X designated samples are singleton recombinants,
descending
from recombination nodes that are ancestral only to that sample (10 are XZ; 6
are XE;  3 each from XN and XK; 2 each from XC, XS, XV, XQ and XAB; and 1
sample from each of XB, XM, XJ, XAF, XAH and XAJ). Such samples are likely to
be enriched for sequencing errors and lineage designation artefacts, and so we
exclude them from further analysis in this section.

The most recent recombination node for 79 samples (XN: 53, XZ: 16, XAJ: 6, XE:
1, XAD: 1, XAH: 1, XAK: 1) is the same. This node, caused by strain
Denmark/DCGC-281594/2021, is ancestral to more than 127K samples,
and is highly likely to be a spuriously inferred recombination event.
Understanding the origin of this
and other similar high-impact, but likely spurious recombination
events is an important aspect of future work.
For simplicity, we exclude these samples from further analyses in
this section.

The remaining remaining 630 Pango recombinant samples trace back to 50 different
most recent recombination nodes, corresponding to 31
Nextclade-designated Pango lineages. These are summarised
in Supplementary Table~\ref{tab:pango-recombinants}, and fall
into three classes: single origin, multiple origin,
and multiple nested origins, which we discuss in the following sections.

@@

\caption{\label{fig:pango-simple-origin-graph} Non-nested
Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by
\texttt{sc2ts}, with
imputed Pango status. Genomic regions inherited by the recombination node are shown
corresponding to the maximum inferred breakpoint position from the HMM process.
(B) Equivalent plot for the 17 XAG samples in the long ARG, with abbreviated labelling.
(C) Equivalent plot for both origination events involving the four XD samples in the long ARG. Details of the mutations and sample node identities for all three plots are provided in supplementary Figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}.
} \end{figure}

@@

It is clear from the XA subgraph that all the samples labelled XA by Nextclade trace to a
single originating recombination node, whose genome is a composite of a B.1.177.18 lineage
on the left of the genome and a B.1.1.7 lineage on the right. In the subgraph we show
the maximum genomic position for the recombination breakpoint, here at 22,227 basepairs
(corresponding to a breakpoint strictly less than 22,228, see Table~\ref{tab:jackson})

@@

\subsubsection{Wide ARG}
\label{sec-pango_x_wide_arg}.
%% Quick overview of the wide ARG.
Since the Wide ARG is restricted to data collected prior
to mid-2021, it contains samples from only three Pango-designated recombinant
lineages: XA, XB, and XC. Both Nextclade and GISAID designate 44
samples as XA, while 237 samples are designated
as XB by Nextclade (231 by GISAID), and 6 as XC by Nextclade (none by GISAID).
After removing singleton recombinants, XA numbers remain unchanged, but XB is
reduced to 235 Nextclade-designated samples (229 GISAID) and XC is reduced to 4
(none in the GISAID designations). We correctly identify that all samples
designated as XA, XB, or XC by any method have one or more recombination nodes
in their ancestry.
